---
title             : 'tidyTouch: An interactive visualization tool for data science education'
shorttitle        : "tidyTouch"

author: 
  - name          : "Jonah DeVaney"
    affiliation   : "1"
  - name          : "Matthew McBee"
    affiliation   : "1"
    
affiliation:
  - id            : "1"
    institution   : "East Tennesse State University"

abstract: |
  One or two sentences providing a **basic introduction** to the field,  comprehensible to a scientist in any discipline.
  
  Two to three sentences of **more detailed background**, comprehensible  to scientists in related disciplines.
  
  One sentence clearly stating the **general problem** being addressed by  this particular study.
  
  One sentence summarizing the main result (with the words "**here we show**" or their equivalent).
  
  Two or three sentences explaining what the **main result** reveals in direct comparison to what was thought to be the case previously, or how the  main result adds to previous knowledge.
  
  One or two sentences to put the results into a more **general context**.
  
  Two or three sentences to provide a **broader perspective**, readily comprehensible to a scientist in any discipline.
  
  <!-- https://tinyurl.com/ybremelq -->
  
keywords          : ""
wordcount         : ""

bibliography      : 
  - ["r-references.bib"]
  - ["add_refs.bib"]

floatsintext      : no
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library("papaja")
library("ggplot2")
library("dplyr")
library("shiny")
library("shinymeta")
library("shinythemes")
library("readr")
library("readxl")
library("tidyr")
library("datasets")
library("haven")
library("reactable")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

# Introduction
intro to FOSS, R, data science education, 
* Data Visualization




# -------------------------Break
# Methods
We report how we determined our sample size, all data exclusions (if any), all manipulations, and all measures in the study. <!-- 21-word solution (Simmons, Nelson & Simonsohn, 2012; retrieved from http://ssrn.com/abstract=2160588) -->

## Participants

## Material

## Procedure

## Data analysis

# Results

# Discussion

## R Packages and Session Info
To recognize those that contribute to R, tools used by members of the R community, and a continually developing field of data science, the software used in creating the tidyTouch app is listed: `r cite_r("r-references.bib")`. 
The session info for this project in its current state - containing the R version used for development and additional loaded packages - is listed below.
```{r echo=FALSE}
sessionInfo()
```


\newpage

# References
```{r create_r-references, message=FALSE, warning=FALSE}
r_refs(file = "r-references.bib")
```

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id="refs" custom-style="Bibliography"></div>
\endgroup
